Computational biology in the 21st century: Algorithms that scale

### 2015-07-29 main.tex
supplement we mightn't need.

When a reference sequence is available, NGS reads can be \emph{mapped} onto
this reference (Figure~\ref{fig:ngs-pipeline}).
Mapping allows the differences between the newly-sequenced genome and the 
reference to be analyzed; these differences, or variants, may include single-nucleotide polymorphisms
(SNPs, which are the genetic analogue to bit-flips) (Figure~\ref{fig:snps}), 
insertions or deletions, or larger-scale changes in the genome.
Determining the differences between an individual genome and a reference is
known as \emph{variant calling}.
While reference-based read mapping is a fundamentally simpler problem than
\emph{de novo} assembly, it is still computationally complex, as gigabytes or
terabytes of reads must each be mapped onto the reference genome, which can 
range from millions (for bacteria) to billions (for mammals) of base pairs.
As an example, the ICGC-TCGA Pan Cancer Analysis of Whole Genomes 
(PCAWG)~\cite{weinstein2013cancer} brings together more than 500 top cancer researchers from about 80 institutions in a coordinated manner with the goal of mapping the entire mutational landscape of 37 common cancer types. 
Currently each sample requires seven hours to download. 
% YWY: 7 hours for whom? Perhaps "even on an institutional connection"?
Importantly, researchers do not trust the provided mapping, and thus they redo mappings. 
The time spent on mapping is about 50\% of the overall time spent on the sequence analysis pipeline. 
As read mapping is typically the most costly step in NGS analysis pipelines (e.g., GATK), any improvement to existing mappers will immediately accelerate sequence analysis studies on large read datasets.

\begin{figure*}[htb!]
\centering
\subfloat[`Shotgun' sequencing breaks DNA molecules into many short fragments, which are read from one or both ends in the form of {\em reads,\/} and relies on high coverage to produce a statistically likely representation of a whole genome]{
\includegraphics[height=2.6in]{assets/shotgun.png}
\label{fig:shotgun}
}
\subfloat[Single-nucleotide polymorphisms, or SNPs, are the simplest type of genomic variant, and form the bulk of `variant-calling' analysis]{
\includegraphics[height=2.6in]{assets/snps.png}
\label{fig:snps}
}\\
\subfloat[The NGS downstream analysis pipeline. Shotgun reads are mapped to a reference genome with tools such as BWA or Bowtie. The resulting genomic sequence is analyzed for variants with tools such as GATK or Samtools. This allows relationships between genes and diseases to be uncovered.]{
\includegraphics[width=4.0in]{assets/ngs-pipeline.png}
\label{fig:ngs-pipeline}
}
\caption[The NGS Pipeline]{The next-generation sequencing (NGS) pipeline.}
\label{fig:ngs}
\end{figure*}

@@

On a potentially even larger scale is the growth of metagenomic data.
Metagenomics is the study of the many genomes (bacterial, fungal, and even 
viral) that make up a particular environment.
Such an environment could be soil from a particular region (which can lead to 
the discovery of new 
antibiotics~\cite{gillespie2002isolation,forsberg2012shared}), or it could be 
the human gut, whose microbiome has been linked to human-health concerns 
including Autism Spectrum Disorder~\cite{macfabe2012short}, 
Crohn's Disease~\cite{gevers2014treatment,manichanh2006reduced}, and 
obesity~\cite{greenblum2012metagenomic}.
Metagenomics fundamentally asks what organisms are present, and, in the case
of a microbiome such as the gut, what metabolic functions it can accomplish as
a whole.
One way of addressing this problem is to attempt to map NGS reads from a metagenomic sample
onto a set of reference genomes that are expected to be present.
This is exactly the read-mapping problem discussed above, but with many 
reference genomes, compounding the computational requirements.
A second way is to perform homology search on a protein sequence database;
exact or nearly-exact matches imply the presence of a species, while more 
distant hits may still give clues to function.
For this task, BLASTX~\cite{altschul1990basic} is commonly
used to translate nucleotide reads into their possible protein sequences, and
search for them in a protein database.
% YWY: 16S isn't well introduced
The difficulty is that the datasets required to shine any light on these 
questions, namely from `shotgun' metagenomics, are gigantic and vastly more 
complex than the 16S metagenomic studies that merely classify the bacteria 
based on 
their divergence with respect to a specific gene (namely, the 16S ribosomal 
subunit).
The massive data results in major identification 
challenges for certain bacterial, as well as viral, species and 
genera~\cite{janda200716s}.

@@

Over the course of evolutionary time, structure is known to be more highly 
conserved than sequence~\cite{illergaard2009structure}, which is to say that it 
does not change as rapidly.
When the structure of a protein is known but its biological function or 
evolutionary relationships are not, researchers may search for structurally
similar proteins that are better studied~\cite{gibrat1996surprising}.
Classical tools for this involve performing pairwise structural alignments to
look for geometric similarity; DALI~\cite{holm1995dali} is still widely used,
along with other aligners such as FATCAT~\cite{ye2004fatcat} and 
Matt~\cite{menke2008matt}.
Due to the complexity of protein structures, these programs generally take 
significant amounts of time,
especially to align multiple structures.
For example, the DALI webserver can take as much as an hour to return results
for a single query.
FragBag~\cite{budowski2010fragbag} accelerates protein structure search by
approximating structural alignments
by instead comparing the `bag-of-words' from each structure.
Analogous to a term-frequency vector in information retrieval, this bag-of-words
indicates the abundance of particular, short structural motifs within a protein.

@@

% In the realm of protein structure prediction, it has been
% observed~\cite{dunbrack2002rotamer} that rather than exploring the continuous
% range of
% possible dihedral angles, protein structures inhabit a smaller subspace, defined
% by a narrower range of angles, referred to as
% rotamers~\cite{dunbrack1993backbone}.
% The TreePack algorithm~\cite{xu2006fast} takes advantage of this observation
% for the
% purpose of sidechain packing (finding the lowest-energy conformation for the
% sidechain of each amino acid), using tree width to reduce the search space and
% simplify dependencies.

>>>>

When a reference sequence is available, NGS reads can be \emph{mapped} onto
this reference (Figure~\ref{fig:ngs-pipeline}).
Mapping allows the differences between the newly-sequenced genome and the 
reference to be analyzed; these differences, or variants, may include single-nucleotide polymorphisms
(SNPs, which are the genetic analogue to bit-flips) (Figure~\ref{fig:snps}), 
insertions or deletions, or larger-scale changes in the genome.
Determining the differences between an individual genome and a reference is
known as \emph{variant calling}.
While reference-based read mapping is a fundamentally simpler problem than
\emph{de novo} assembly, it is still computationally complex, as gigabytes or
terabytes of reads must each be mapped onto the reference genome, which can 
range from millions (for bacteria) to billions (for mammals) of base pairs.
As an example, the ICGC-TCGA Pan Cancer Analysis of Whole Genomes 
(PCAWG)~\cite{weinstein2013cancer} brings together more than 500 top cancer 
researchers from about 80 institutions in a coordinated manner with the goal of 
mapping the entire mutational landscape of 37 common cancer types. 
Currently each sample requires seven hours to download even on an institutional 
connection. 
Importantly, researchers do not trust the provided mapping, and thus they redo mappings. 
The time spent on mapping is about 50\% of the overall time spent on the sequence analysis pipeline. 
As read mapping is typically the most costly step in NGS analysis pipelines (e.g., GATK), any improvement to existing mappers will immediately accelerate sequence analysis studies on large read datasets.

\begin{figure*}[htb!]
\centering
\subfloat[`Shotgun' sequencing breaks DNA molecules into many short fragments, which are read from one or both ends in the form of {\em reads,\/} and relies on high coverage to produce a statistically likely representation of a whole genome]{
\includegraphics[height=2.0in]{assets/shotgun.png}
\label{fig:shotgun}
}
\subfloat[Single-nucleotide polymorphisms, or SNPs, are the simplest type of genomic variant, and form the bulk of `variant-calling' analysis]{
\includegraphics[height=2.0in]{assets/snps.png}
\label{fig:snps}
}\\
\subfloat[The NGS downstream analysis pipeline. Shotgun reads are mapped to a reference genome with tools such as BWA or Bowtie. The resulting genomic sequence is analyzed for variants with tools such as GATK or Samtools. This allows relationships between genes and diseases to be uncovered.]{
\includegraphics[width=5.0in]{assets/ngs-pipeline.pdf}
\label{fig:ngs-pipeline}
}
\caption[The NGS Pipeline]{The next-generation sequencing (NGS) pipeline.}
\label{fig:ngs}
\end{figure*}

@@

On a potentially even larger scale is the growth of metagenomic data.
Metagenomics is the study of the many genomes (bacterial, fungal, and even 
viral) that make up a particular environment.
Such an environment could be soil from a particular region (which can lead to 
the discovery of new 
antibiotics~\cite{gillespie2002isolation,forsberg2012shared}), or it could be 
the human gut, whose microbiome has been linked to human-health concerns 
including Autism Spectrum Disorder~\cite{macfabe2012short}, 
Crohn's Disease~\cite{gevers2014treatment,manichanh2006reduced}, and 
obesity~\cite{greenblum2012metagenomic}.
Metagenomics fundamentally asks what organisms are present, and, in the case
of a microbiome such as the gut, what metabolic functions it can accomplish as
a whole.
One way of addressing this problem is to attempt to map NGS reads from a metagenomic sample
onto a set of reference genomes that are expected to be present.
This is exactly the read-mapping problem discussed above, but with many 
reference genomes, compounding the computational requirements.
A second way is to perform homology search on a protein sequence database;
exact or nearly-exact matches imply the presence of a species, while more 
distant hits may still give clues to function.
For this task, BLASTX~\cite{altschul1990basic} is commonly
used to translate nucleotide reads into their possible protein sequences, and
search for them in a protein database.
The difficulty is that the datasets required to shine any light on these 
questions, namely from `shotgun' metagenomics, are gigantic and vastly more 
complex than the 16S metagenomic studies that merely classify the bacteria 
based on 
their divergence with respect to a specific gene (namely, the 16S ribosomal 
subunit).
The massive data results in major identification 
challenges for certain bacterial, as well as viral, species and 
genera~\cite{janda200716s}.

@@

@@
